---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
library(knitr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
knit_print.data.frame = function(x, ...) {
    res = paste(c("", "", knitr::kable(x, output = FALSE)), collapse = "\n")
    asis_output(res)
}
# register the method
registerS3method("knit_print", "data.frame", knit_print.data.frame)
library("FUNGuildR")
```
# FUNGuildR

<!-- badges: start -->
[![Travis build status](https://travis-ci.com/brendanf/FUNGuildR.svg?branch=master)](https://travis-ci.com/brendanf/FUNGuildR)
[![Codecov test coverage](https://codecov.io/gh/brendanf/FUNGuildR/branch/master/graph/badge.svg)](https://codecov.io/gh/brendanf/FUNGuildR?branch=master)
<!-- badges: end -->

This is a simple reimplementation of FUNGuild_v1.1.py. It queries
the [FUNGuild or NEMAGuild databases](http://www.stbates.org/guilds/app.php)
and assigns trait information based on matching to a taxonomic
classification. It does not include a copy of the FUNGuild or NEMAGuild
databases, because they are continually updated as new information is
submitted, but it does have methods to download them and store them as R
objects to speed up repeated queries or to allow local queries without
internet access.

## Installation

For the moment, `FUNGuildR` can only be installed from GitHub.
To install, be sure you have installed the [devtools](https://cran.r-project.org/web/packages/devtools/index.html) package, and then type

``` r
devtools::install_github("brendanf/FUNGuildR")
```

## Usage

The main functions are `funguild_assign()` and `nemaguild_assign()`.
Each of these takes as their first argument a `data.frame`, which should contain a `character` column named "Taxonomy".
They return a version of the same `data.frame` (as a `tibble`) with additional columns "taxon", "taxonomicLevel", "trophicMode", "guild", "growthForm", "trait", "confidenceRanking", "notes", and "citationSource".
That's it!

Here's how it works on a sample database (scroll to see additional columns):
```{r}
sample_fungi

sample_guilds <- funguild_assign(sample_fungi)
sample_guilds
```

For more information about the meaning of the new columns, see the FUNGuild manual, available at https://github.com/UMNFuN/FUNGuild.

## Import data format
Each value in the Taxonomy column should consist of a comma-, colon-, underscore_, or semicolon-delimited list of taxa which the organism on that row belongs to.
You can see examples in the `sample_fungi` data presented above.

Such taxonomic classifications are frequently arranged from the most inclusive taxon (e.g., Kingdom) to the most specific (e.g., Species), but this is not actually required.
Not all taxonomic ranks are required; for each row, the algorithm returns results only for the most specific taxon which is present in the database.

## Database caching

By default, `funguild_assign()` and `nemaguild_assign()` download the appropriate database each time they are invoked.
In many analysis pipelines, where guilds need to be assigned only once, this is not a problem; because the databases are continuously updated, it is good to use the most current version.
However, if you are going to call the functions many times, or if you plan to assign guilds in a situation where you have no internet access, you can cache the database(s) locally using the functions `get_funguild_db()` and `get_nemaguild_db()`.
Each of these returns the relevant database as a `tibble`, which can be passed as the second argument to `funguild_assign()` or `nemaguild_assign()`.

```r
nema <- get_nemaguild_db()

# This isn't necessary for a single query, but it works.
nema_guild <- nemaguild_assign(sample_nema, db = nema)

# It might be more useful in this situation
data_guilds <- lapply(many_datasets, nemaguild_assign, db = nema)

# Or you can save it for later offline use
saveRDS(nema, "nemaguild.rds")

#And then load it again
nema <- loadRDS("nemaguild.rds")
```

This strategy can also be used for reproduceable research, to store the same version of the database which was used in the original paper.
